---
title: Risk of developing a symptomatic or severe form of COVID-19 depending on the
  clinical form of the infecting individual
author: "Thomas Kesteman"
date: "14/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load myData, include=FALSE}
load("C:/Users/thoma/Documents/3.7 COVID-19/Modelling - Marc/Rmarkdown/201014 both datasets.RData") 
```

# Description of data
### Dataset 1
Line list Vietnam “Pos COVID-19 413 update” + contact informations from Tan Le Van for 2 individuals from the “Buddha bar” cluster (data received 08/05/2020)

N individuals in the dataset: `r sum(llr$wave==1)`

### Dataset 2
Line list Vietnam Da Nang (rds file as of 16 Sep 2020)

N individuals in the dataset: `r sum(llr$wave==2)`

## Contacts
N individuals whom contact(s) has been identified: `r sum(llr$n_contacts>0)`

Out of the `r sum(llr$n_contacts_before_clean %in% 1)` individuals with 1 contact mentioned, there were `r sum(llr$pair_infectorinfectee_YN)/2 + sum(llr$n_contacts_before_clean %in% 1 & llr$n_contacts %in% 0)` pairs of cases whom the infector was the infectee and vice-versa. Using dates of onset, the infector could be identified in `r sum(llr$n_contacts_before_clean %in% 1 & llr$n_contacts %in% 0)` pairs; in the other pairs, there was no possibility to know who was the infector and who was the infectee.

```{r results = 'asis', echo=FALSE}
mytable = with(llr,table(n_contacts,paste("Dataset",wave)))
library(kableExtra)
knitr::kable(mytable, "html", caption = "N of contacts identified per dataset") %>%
  kable_styling(full_width = F,position = "left")
```

## Comorbidities
`r sum(llr$Diabetes)` cases had diabetes, `r sum(llr$Cancer)` cases had cancer, `r sum(llr$Obesity, na.rm=T)` had obesity, `r sum(llr$Hypertension)` had hypertension, `r sum(llr$Kidney_failure, na.rm=T)` had kidney failure, and `r sum(llr$Cardiovascular | llr$Cardiac_injury, na.rm=T)` had other cardiovascular pathologies. Comorbidities such as arthritis, asthma, epilepsy, vestibular disorders, congenital hemolysis, gastritis, spinal pain, urinary tract infections, and sinusitis have been disregarded as regards COVID-19. Overall, `r sum(llr$Comorbidity)` cases had a significant comorbidity as regards COVID-19.

## Clinical forms of COVID-19 
### Symptomatic vs asymptomatic
The definition of asymptomatic didn't vary much between the 2 datasets. Therefore, results can be compared.

* In the whole dataset, there were `r sum(llr$Symptomatic %in% F)` asymptomatic cases (`r format(100*sum(llr$Symptomatic %in% F)/nrow(llr), digits=1, nsmall=1)`%), `r sum(llr$Symptomatic %in% T)` symptomatic (`r format(100*sum(llr$Symptomatic %in% T)/nrow(llr), digits=1, nsmall=1)`%), and `r sum(llr$Symptomatic %in% NA)` NA (`r format(100*sum(llr$Symptomatic %in% NA)/nrow(llr), digits=1, nsmall=1)`%). 

* Out of the `r sum(llr$n_contacts>0)` cases whose contact has been identified, there were `r sum(llr$Symptomatic %in% F & llr$n_contacts>0)` asymptomatic cases (`r format(100*sum(llr$Symptomatic %in% F & llr$n_contacts>0)/sum(llr$n_contacts>0), digits=1, nsmall=1)`%), `r sum(llr$Symptomatic %in% T & llr$n_contacts>0)` symptomatic (`r format(100*sum(llr$Symptomatic %in% T & llr$n_contacts>0)/sum(llr$n_contacts>0), digits=1, nsmall=1)`%), and `r sum(llr$Symptomatic %in% NA & llr$n_contacts>0)` NA (`r format(100*sum(llr$Symptomatic %in% NA & llr$n_contacts>0)/sum(llr$n_contacts>0), digits=1, nsmall=1)`%). 

  + In dataset 1, there were `r sum(llr$n_contacts>0 & llr$wave==1)` cases whose contact has been identified, including `r sum(llr$Symptomatic %in% F & llr$n_contacts>0 & llr$wave==1)` asymptomatic cases (`r format(100*sum(llr$Symptomatic %in% F & llr$n_contacts>0 & llr$wave==1)/sum(llr$n_contacts>0 & llr$wave==1), digits=1, nsmall=1)`%), `r sum(llr$Symptomatic %in% T & llr$n_contacts>0 & llr$wave==1)` symptomatic (`r format(100*sum(llr$Symptomatic %in% T & llr$n_contacts>0 & llr$wave==1)/sum(llr$n_contacts>0 & llr$wave==1), digits=1, nsmall=1)`%), and `r sum(llr$Symptomatic %in% NA & llr$n_contacts>0 & llr$wave==1)` NA (`r format(100*sum(llr$Symptomatic %in% NA & llr$n_contacts>0 & llr$wave==1)/sum(llr$n_contacts>0 & llr$wave==1), digits=1, nsmall=1)`%). 
  
  + In dataset 2, there were `r sum(llr$n_contacts>0 & llr$wave==2)` cases whose contact has been identified, including `r sum(llr$Symptomatic %in% F & llr$n_contacts>0 & llr$wave==2)` asymptomatic cases (`r format(100*sum(llr$Symptomatic %in% F & llr$n_contacts>0 & llr$wave==2)/sum(llr$n_contacts>0 & llr$wave==2), digits=1, nsmall=1)`%), `r sum(llr$Symptomatic %in% T & llr$n_contacts>0 & llr$wave==2)` symptomatic (`r format(100*sum(llr$Symptomatic %in% T & llr$n_contacts>0 & llr$wave==2)/sum(llr$n_contacts>0 & llr$wave==2), digits=1, nsmall=1)`%), and `r sum(llr$Symptomatic %in% NA & llr$n_contacts>0 & llr$wave==2)` NA (`r format(100*sum(llr$Symptomatic %in% NA & llr$n_contacts>0 & llr$wave==2)/sum(llr$n_contacts>0 & llr$wave==2), digits=1, nsmall=1)`%). 

  + Proportions of asymptomatic cases have been inversed (`r format(100*sum(llr$Symptomatic %in% F & llr$n_contacts>0 & llr$wave==1)/sum(llr$n_contacts>0 & llr$wave==1), digits=1, nsmall=1)`% vs `r format(100*sum(llr$Symptomatic %in% F & llr$n_contacts>0 & llr$wave==2)/sum(llr$n_contacts>0 & llr$wave==2), digits=1, nsmall=1)`%, p = `r format(with(llr, fisher.test(Symptomatic,wave))$p.value, digits=2, nsmall=1)`) between the 2 datasets, despite a similar definition.


### Severe vs non-severe
The definition of severity changed between the 2 datasets: in dataset 1, it's defined as any need in oxygen, while in dataset 2 it was defined as "severity" (defined by the clinician) at admission or fatal outcome. Therefore, results cannot be compared.

* In dataset 1, there were `r sum(llr$Severe %in% T & llr$wave==1)` severe cases (`r format(100*sum(llr$Severe %in% T & llr$wave==1)/sum(llr$wave==1), digits=1, nsmall=1)`%), and `r sum(llr$Severe %in% NA & llr$wave==1)` NA (symptom NA were considered as non-severe).

  + In this dataset, among `r sum(llr$n_contacts>0 & llr$wave==1)` cases whose contact has been identified, there were `r sum(llr$Severe %in% T & llr$n_contacts>0 & llr$wave==1)` severe cases (`r format(100*sum(llr$Severe %in% T & llr$n_contacts>0 & llr$wave==1)/sum(llr$n_contacts>0 & llr$wave==1), digits=1, nsmall=1)`%). 
  
* In dataset 2, there were `r sum(llr$Severe %in% T & llr$wave==2)` severe cases (`r format(100*sum(llr$Severe %in% T & llr$wave==2)/sum(llr$wave==2), digits=1, nsmall=1)`%).

  + In this dataset, among `r sum(llr$n_contacts>0 & llr$wave==2)` cases whose contact has been identified, there were `r sum(llr$Severe %in% T & llr$n_contacts>0 & llr$wave==2)` severe cases (`r format(100*sum(llr$Severe %in% T & llr$n_contacts>0 & llr$wave==2)/sum(llr$n_contacts>0 & llr$wave==2), digits=1, nsmall=1)`%). 

* Proportions of severe cases were similar (`r format(100*sum(llr$Severe %in% T & llr$wave==1)/sum(llr$wave==1), digits=1, nsmall=1)`% vs `r format(100*sum(llr$Severe %in% T & llr$wave==2)/sum(llr$wave==2), digits=1, nsmall=1)`%, p = `r format(with(llr, fisher.test(Severe,wave))$p.value, digits=2, nsmall=1)`) between the 2 datasets, despite a different definition.

# Clinical form of infectee as a function of the clinical form of infector
## Effect of a symptomatic infector
### Association between severity and symptomatic infector
```{r, echo=FALSE}
t = with(llr,table(Severe,contact_with_symptomatic))
```

The probability to develop a severe form is higher in contacts of a symptomatic case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`%) than in contacts of an asymptomatic case (`r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), but this difference is non-significant (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). 


```{r, echo=FALSE,include=F}
library(gee)
geeORtable = function(fit,ref_txt = "1.00", sepCI=" - ",interact=F) { 
  fit_res = as.data.frame(paste0(format(round(exp(fit$coef[-1]),2), nsmall=2)," [",
                                 sub(" ","",format(round(exp(coef(fit)-summary(fit)$coef[,4]*qnorm(0.975)),2), nsmall=2)[-1]),sepCI,
                                 sub(" ","",format(round(exp(coef(fit)+summary(fit)$coef[,4]*qnorm(0.975)),2), nsmall=2)[-1]),"]"))
  colnames(fit_res) = "ORCI"
  fit_res = cbind(fit_res,pval =  as.character(ifelse(2*pnorm(-abs(summary(fit)$coef[-1,5]))<0.001,"<0.001",round(2*pnorm(-abs(summary(fit)$coef[-1,5])),3))))
  fit_res = cbind(fit_res,Sign = cut(2*pnorm(-abs(summary(fit)$coef[-1,5])),breaks=c(0,0.001,0.01,0.05,1),labels=c("***","**","*",""),include.lowest=T))
  fit_res = cbind(Categ =labels(fit$coef[-1]), fit_res)
  if (!interact) {
    varfit = gsub(" ","",unlist(strsplit(substr(as.character(fit$call)[2],regexpr("~",as.character(fit$call)[2])[1]+2,nchar(as.character(fit$call)[2])),"[+]")))
    labels_varfit = lapply(varfit,FUN=function(x) {eval(parse(text=paste0("levels(",as.character(fit$call)[4],"$",x,")")))})
    varfit_rep = rep(varfit,unlist(lapply(labels_varfit,FUN=function(x){length(x)})))
    Categ = paste0(varfit_rep,unlist(labels_varfit))
    order_var= 	rep(c(1:length(varfit)),unlist(lapply(labels_varfit,FUN=function(x){length(x)})))
    tabnames_varfit = cbind(as.data.frame(Categ),as.data.frame(order_var))
    fit_res = merge(tabnames_varfit, fit_res,all=T)
    colnames(fit_res) = c("Categ.","order_var","Adjusted OR [95%IC]","pval","Signif.")
    fit_res[,5] = ifelse(is.na(fit_res[,5]),"",as.character(fit_res[,5]))
    fit_res[,4] = ifelse(is.na(fit_res[,4]),"",as.character(fit_res[,4]))
    fit_res[,3] = ifelse(is.na(fit_res[,3]),ref_txt,as.character(fit_res[,3]))
    fit_res  = fit_res[order(fit_res$order_var, fit_res$Categ.),]
    fit_res  = fit_res[,-2]
  }
  return(fit_res) 
}
fit = gee(Severe ~ contact_with_symptomatic + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,corstr="exchangeable") 
```
Nevertheless, in a multivariate analysis (GEE), taking into account the age, the gender, and the comorbidities of the individual and considering an exchangeable within-dataset correlation structure, the relationship between severity and being contact of a symptomatic case becomes significant (OR `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```


```{r, echo=FALSE}
t = with(llr,table(Severe,contact_with_sympt_attimeofcontact))
```
```{r, echo=FALSE,include=F}
fit = gee(Severe ~ contact_with_sympt_attimeofcontact + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,corstr="exchangeable") 
```

Similarly, if symptomatic infectors are defined as those who had developed symptoms at time of contact with the infectee, the probability to develop a severe form is higher in contacts of a symptomatic case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`%) than in contacts of an asymptomatic case (`r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), but this difference is non-significant (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). And in a multivariate analysis, the relationship between severity and being contact of a symptomatic case becomes significant as well (OR `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```

### Association between symptomatic form and symptomatic infector
```{r, echo=FALSE}
t = with(llr,table(Symptomatic,contact_with_symptomatic))
```

The probability to develop a symptomatic form is higher in contacts of a symptomatic case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`%) than in contacts of an asymptomatic case (`r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), and this difference is significant (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). 


```{r, echo=FALSE,include=F}
fit = gee(Symptomatic ~ contact_with_symptomatic + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,corstr="exchangeable") 
```
In a multivariate analysis (GEE), taking into account the age, the gender, and the comorbidities of the individual and considering an exchangeable within-dataset correlation structure, the relationship between symptomatic form and being contact of a symptomatic case remains significant (OR `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```


```{r, echo=FALSE}
t = with(llr,table(Symptomatic,contact_with_sympt_attimeofcontact))
```
```{r, echo=FALSE,include=F}
fit = gee(Symptomatic ~ contact_with_sympt_attimeofcontact + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,corstr="exchangeable") 
```

Similarly, if symptomatic infectors are defined as those who had developed symptoms at time of contact with the infectee, the probability to develop a symptomatic form is higher in contacts of a symptomatic case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`%) than in contacts of an asymptomatic case (`r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), and this difference is significant but less marked than with the previous definition of contact with symptomatic case (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). And in a multivariate analysis, the relationship between symptomatic form and being contact of a symptomatic case remains significant as well (OR `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```



```{r, echo=FALSE}
t = with(llr[!llr$pair_infectorinfectee_YN,],table(Symptomatic,contact_with_symptomatic))
```
```{r, echo=FALSE,include=F}
fit = gee(Symptomatic ~ contact_with_symptomatic + Age + Gender_FM + Comorbidity,
          data=llr[!llr$pair_infectorinfectee_YN,],id=wave,family=binomial,corstr="exchangeable") 
```

When excluding the pairs where infector is infectee, the probability to develop a symptomatic form remains higher in contacts of a symptomatic case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`%) than in contacts of an asymptomatic case (`r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), and this difference is significant (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). And in a multivariate analysis, the relationship between symptomatic form and being contact of a symptomatic case remains significant as well (OR `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```


## Effect of an infector with a severe form
### Association between severity in infectee and severity in infector
```{r, echo=FALSE}
t = with(llr,table(Severe,contact_with_severe))
```

The probability to develop a severe form is higher in contacts of a severe case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`%) than in contacts of an asymptomatic case or non-severe case (`r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), and this difference is significant (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). 


```{r, echo=FALSE,include=F}
fit = gee(Severe ~ contact_with_severe + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,corstr="exchangeable") 
```
In a multivariate analysis (GEE), taking into account the age, the gender, and the comorbidities of the individual and considering an exchangeable within-dataset correlation structure, the relationship between severity and being contact of a severe case remains significant (OR `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```


```{r, echo=FALSE}
llr$contact_w_severe_sympt_atcontact = llr$contact_with_severe_sympt_attimeofcontact
t = with(llr,table(Severe,contact_w_severe_sympt_atcontact))
```
```{r, echo=FALSE,include=F}
fit = gee(Severe ~ contact_w_severe_sympt_atcontact + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,corstr="exchangeable") 
```

Similarly, if we restrict the analysis to those infectors who had developed symptoms before contact with the infectee, the probability to develop a severe form is higher in contacts of a severe case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`% vs `r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), and this difference is significant (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). In a multivariate analysis, this relationship remains significant as well (OR `r geeORtable(fit)[grepl("contact_w", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_w", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```

```{r, echo=FALSE}
t = with(llr[!llr$pair_infectorinfectee_YN,],table(Severe,contact_with_severe))
```
```{r, echo=FALSE,include=F}
fit = gee(Severe ~ contact_with_severe + Age + Gender_FM + Comorbidity,
          data=llr[!llr$pair_infectorinfectee_YN,],id=wave,family=binomial,corstr="exchangeable") 
```

When excluding the pairs where infector is infectee and vice-versa, the probability to develop a severe form remains higher in contacts of a severe case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`% vs `r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), and this difference is significant (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). In a multivariate analysis, this relationship remains significant as well (OR `r geeORtable(fit)[grepl("contact_w", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_w", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```

### Association between symptomatic form and severity in infector
```{r, echo=FALSE}
t = with(llr,table(Symptomatic,contact_with_severe))
```

The probability to develop a symptomatic form is similar in contacts of a severe case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`%) than in contacts of an asymptomatic case or non-severe case (`r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), and this difference is non-significant (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). 

```{r, echo=FALSE,include=F}
fit = gee(Symptomatic ~ contact_with_severe + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,corstr="exchangeable") 
```
In a multivariate analysis (GEE), taking into account the age, the gender, and the comorbidities of the individual and considering an exchangeable within-dataset correlation structure, the relationship between symptomatic form and being contact of a severe case is closer to the significance threshold but remains modest (OR `r geeORtable(fit)[grepl("contact_w", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_w", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```


```{r, echo=FALSE}
t = with(llr,table(Symptomatic,contact_w_severe_sympt_atcontact))
```
```{r, echo=FALSE,include=F}
fit = gee(Symptomatic ~ contact_w_severe_sympt_atcontact + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,corstr="exchangeable") 
```

Similarly, if we restrict the analysis to infectors who had developed symptoms before contact, the probability to develop a symptomatic form is a bit higher in contacts of a severe case (`r t[2,2]`/`r sum(t[,2])`, `r format(prop.table(t,2)[2,2]*100, digits=1, nsmall=1)`%) than in contacts of an asymptomatic case or non-severe case (`r t[2,1]`/`r sum(t[,1])`, `r format(prop.table(t,2)[2,1]*100, digits=1, nsmall=1)`%), and this difference is non-significant in a 2x2 table (OR `r format(fisher.test(t)$estimate, digits=2, nsmall=2)`, Fisher’s exact test p= `r format(fisher.test(t)$p.value, digits=2, nsmall=1)`). In a multivariate analysis taking into account covariables and the population structure, the relationship becomes remains non-significant (OR `r geeORtable(fit)[grepl("contact_w", geeORtable(fit)[,1]),2]`, p `r geeORtable(fit)[grepl("contact_w", geeORtable(fit)[,1]),3]`).
```{r ,echo=F}
geeORtable(fit)
```

# Effect of age of infector

## Age of infectors and infectees

```{r,  echo=FALSE}
age_nocontact = llr$Age[llr$n_contacts==0]
age_infectees = llr$Age[llr$n_contacts>=1]
age_infectors = llr$Age[llr$id %in% unique(unlist(strsplit(as.character(llr$contact_id[!is.na(llr$contact_id)]), "\\,")))]
```
Mean age of infectees (`r format(mean(age_infectees), digits=1, nsmall=1)` years) is similar to mean age of infectors (`r format(mean(age_infectors), digits=1, nsmall=1)` years, Wilcoxon test, p=`r format(wilcox.test(age_infectors,age_infectees)$p.value, digits=1, nsmall=2)`). Both are significantly superior to the mean age of individuals whose infector could not be identified (`r format(mean(age_nocontact), digits=1, nsmall=1)` years, Wilcoxon test, p=`r format(wilcox.test(age_nocontact,age_infectees)$p.value, digits=2, nsmall=2)` and p=`r format(wilcox.test(age_nocontact,age_infectors)$p.value, digits=2, nsmall=2)`, respectively).  

```{r,  echo=FALSE}
library(ggplot2)
library(ggpubr)
df = data.frame(Type=c(rep("No contact",length(age_nocontact)),
                       rep("Infectees",length(age_infectees)),
                       rep("Infectors",length(age_infectors))),
                Age =c(age_nocontact,age_infectees,age_infectors))
my_comparisons <- list( c("No contact", "Infectees"), c("Infectors", "Infectees"), c("No contact", "Infectors"))
g = ggplot(df, aes(x = factor(Type), y = Age))  +
  geom_violin() + geom_boxplot(width=0.1) +
  ylab("Age") +
  scale_y_continuous(breaks=seq(0, 110+15, 10)) +
  theme(legend.position = "none",axis.title.x=element_blank(),axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank())
g + stat_compare_means(comparisons = my_comparisons, label.y = c(90,95,100)+15, label = "p.signif") +
  stat_compare_means(label.y = 120)
```

```{r,  echo=FALSE}
correl <- with(llr[!is.na(llr$age_contact),],cor.test(Age, age_contact))
```

The age of infectors is poorly correlated with the age of infectees (Spearman r=`r format(correl$est, digits=3, nsmall=1)`, p=`r format(correl$p.value, digits=2, nsmall=1)`)

```{r,  echo=FALSE}
df=as.data.frame(with(llr,table(age_gp,age_contact_gp)))
maxcount = max(df$Freq)
df$color = ifelse(df$Freq<1,"0","1")
ggplot(df, aes(x= age_contact_gp, y= age_gp)) + geom_point(aes(size = Freq,color=color),na.rm = TRUE) +
  xlab("Age of infector") + ylab("Age of infectee") + scale_size(range = c(1, 10),name="Count") +
  scale_color_manual(values = c("white", "black")) + guides(color=FALSE)
```

## Effect of adding the age of contact in the model
### Influence of infector's age on the association between symptomatic form and symptomatic infector
```{r, echo=FALSE,include=F}
fit_ref = gee(Symptomatic ~ contact_with_symptomatic + Age + Gender_FM + Comorbidity,
          data=llr[!is.na(llr$age_contact),],id=wave,family=binomial,
          corstr="exchangeable") 

fit = gee(Symptomatic ~ contact_with_symptomatic + age_contact + Age + Gender_FM + Comorbidity,
          data=llr[!is.na(llr$age_contact),],id=wave,family=binomial,
          corstr="exchangeable") 
```

In a multivariate analysis, the age of the infector slightly **decreases** the risk of developing a symptomatic form (OR `r geeORtable(fit)[grepl("age_contact", geeORtable(fit)[,1]),2]` per age year) and this relationship is close to significance threshold (p= `r geeORtable(fit)[grepl("age_contact", geeORtable(fit)[,1]),3]`). Adding the age of the infector modifies slightly the relationship between symptomatic form and symptomatic infector: in the same population, i.e. infectors with 1 contact identified, the OR of being a contact of a symptomatic case was `r geeORtable(fit_ref)[grepl("contact_with", geeORtable(fit_ref)[,1]),2]` when the age of contact was not in the multivariate analysis, and OR was `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),2]` when considering the age of the infector in the model (see table below).

```{r ,echo=F}
geeORtable(fit)
```

### Influence of infector's age on the association between severity in infectee and severity in infector

```{r, echo=FALSE,include=F}
fit_ref = gee(Severe ~ contact_with_severe + Age + Gender_FM + Comorbidity,
          data=llr[!is.na(llr$age_contact),],id=wave,family=binomial,
          corstr="exchangeable") 

fit = gee(Severe ~ contact_with_severe + age_contact + Age + Gender_FM + Comorbidity,
          data=llr[!is.na(llr$age_contact),],id=wave,family=binomial,corstr="exchangeable") 
```

Similarly, the age of the infector **decreases** the risk of developing a severe form (OR `r geeORtable(fit)[grepl("age_contact", geeORtable(fit)[,1]),2]` per age year, p `r geeORtable(fit)[grepl("age_contact", geeORtable(fit)[,1]),3]`). Adding the age of the infector also modifies the relationship between symptomatic form and symptomatic infector: in the same population, i.e. infectors with 1 contact identified, the OR of being a contact of a severe case was `r geeORtable(fit_ref)[grepl("contact_with", geeORtable(fit_ref)[,1]),2]` when the age of contact was not in the multivariate analysis, and OR was `r geeORtable(fit)[grepl("contact_with", geeORtable(fit)[,1]),2]` when considering the age of the infector in the model (see table below).

```{r ,echo=F}
geeORtable(fit)
```

# Influence of healthcare
To investigate whether this observation could be an artifact due to healthcare, e.g. better observation of symptoms among healthcare workers or when hospitalized, we introduced 2 variables in the multivariate model: 

1. healthcare-related profession: there were `r sum(llr$HCW)` healthcare workers in the population (`r format(sum(llr$HCW)*100/nrow(llr), digits=1, nsmall=1)`%);

2. belonging to a hospital cluster: there were `r sum(llr$Hospital_cluster)` individuals whose source of infection was related to an hospital or medical centre in the population (`r format(sum(llr$Hospital_cluster)*100/nrow(llr), digits=1, nsmall=1)`%).

> Lots of individuals in hospital clusters (2/3 in wave 2)! --> to be discussed!!!

### Influence of healthcare on the association between symptomatic form and symptomatic infector
```{r, echo=FALSE,include=F}
fit_ref = gee(Symptomatic ~ contact_with_symptomatic + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 

fit1 = gee(Symptomatic ~ contact_with_symptomatic + HCW + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 

fit2 = gee(Symptomatic ~ contact_with_symptomatic + HCW + Hospital_cluster + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 
```

In a multivariate analysis, being a healthcare workers strongly **decreases** the risk of developing a symptomatic form (OR `r geeORtable(fit1)[grepl("HCW", geeORtable(fit1)[,1]),2]`, p `r geeORtable(fit1)[grepl("HCW", geeORtable(fit1)[,1]),3]`). Nevertheless, adding this variable to the model didn't modify the relationship between symptomatic form and symptomatic infector: the OR of being a contact of a symptomatic case was `r geeORtable(fit_ref)[grepl("contact_with", geeORtable(fit_ref)[,1]),2]` when the variable *healthcare worker* was not in the multivariate analysis, and OR was `r geeORtable(fit1)[grepl("contact_with", geeORtable(fit1)[,1]),2]` when considering the profession.

```{r ,echo=F}
geeORtable(fit1)
```

In contrast, belonging to a healthcare cluster was not associated with the risk of developing a symptomatic form (OR `r geeORtable(fit2)[grepl("Hospital_cluster", geeORtable(fit2)[,1]),2]`, p= `r geeORtable(fit2)[grepl("Hospital_cluster", geeORtable(fit2)[,1]),3]`) and adding this variable didn't modify the relationship between symptomatic form and symptomatic infector: the OR of being a contact of a symptomatic case was `r geeORtable(fit2)[grepl("contact_with", geeORtable(fit2)[,1]),2]` when considering the type of cluster.

```{r ,echo=F}
geeORtable(fit2)
```

### Influence of healthcare on the association between severe form in infector and infectee
```{r, echo=FALSE,include=F}
fit_ref = gee(Severe ~ contact_with_severe + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 

fit1 = gee(Severe ~ contact_with_severe + HCW + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 

fit2 = gee(Severe ~ contact_with_severe + HCW + Hospital_cluster + Age + Gender_FM + Comorbidity,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 
```

In a multivariate analysis, being a healthcare workers strongly **increases** the risk of developing a severe form (OR `r geeORtable(fit1)[grepl("HCW", geeORtable(fit1)[,1]),2]`, p `r geeORtable(fit1)[grepl("HCW", geeORtable(fit1)[,1]),3]`). Nevertheless, adding this variable didn't modify the relationship between severe form in infectee and severe form in infector: the OR of being a contact of a severe case was `r geeORtable(fit_ref)[grepl("contact_with", geeORtable(fit_ref)[,1]),2]` when the variable *healthcare worker* was not in the multivariate analysis, and OR was `r geeORtable(fit1)[grepl("contact_with", geeORtable(fit1)[,1]),2]` when considering the profession.

```{r ,echo=F}
geeORtable(fit1)
```

Belonging to a healthcare cluster was also associated with the risk of developing a severe form (OR `r geeORtable(fit2)[grepl("Hospital_cluster", geeORtable(fit2)[,1]),2]`, p `r geeORtable(fit2)[grepl("Hospital_cluster", geeORtable(fit2)[,1]),3]`), but adding this variable didn't modify (significantly???) the relationship between severe form in infectee and infector: the OR of being a contact of a severe case was `r geeORtable(fit2)[grepl("contact_with", geeORtable(fit2)[,1]),2]` when considering the type of cluster (instead of `r geeORtable(fit_ref)[grepl("contact_with", geeORtable(fit_ref)[,1]),2]`).

> I don't know how to test the significance of changes in a coef (here: `contact_with_severe`) when adding a variable (here: `Hospital_cluster`) to a model.
> `gee(Severe ~ contact_with_severe + Hospital_cluster + HCW + Age + Gender_FM +`
`Comorbidity, data=llr, id=wave, family=binomial, corstr="exchangeable")`

```{r ,echo=F}
geeORtable(fit2)
```


# Superspreaders
## Description
In our population, the majority of cases (`r sum(llr$n_infected==0)`/`r nrow(llr)`, `r format(sum(llr$n_infected==0)*100/nrow(llr), digits=1, nsmall=1)`%) was not listed among the possible infectors, i.e. they possibly didn't infect anyone. A small proportion of cases possibly have been listed in the possible infectors of 5 or more individuals (`r sum(llr$n_infected_categ=="5+")`/`r nrow(llr)`, `r format(sum(llr$n_infected_categ=="5+")*100/nrow(llr), digits=1, nsmall=1)`%). This subgroup will be considered later as *superspreaders*, even if there's no certainty that they were actually the source of infection. Overall, `r format(sum(llr$contact_with_superspreader==0)*100/nrow(llr), digits=1, nsmall=1)`% of cases have had a contact with those superspreaders and, in cases for whom at least one contact is found, this proportion is `r format(sum(llr$contact_with_superspreader[llr$n_contacts>0])*100/ sum(llr$n_contacts>0), digits=1, nsmall=1)`%.

```{r,  echo=FALSE}
ggplot(llr[llr$n_infected!=0,], aes(x=n_infected_max9)) + geom_bar(stat="count") + xlab("N of infectees per infector") + ylab("N of infectors")
```

## Superspreaders and risk of symptomatic form 

```{r, echo=FALSE,include=F}
fit_ref = gee(Symptomatic ~ contact_with_symptomatic + 
                Age + Gender_FM + Comorbidity + HCW + Hospital_cluster,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 

fit1 = gee(Symptomatic ~ contact_with_symptomatic + superspreader +
             Age + Gender_FM + Comorbidity + HCW + Hospital_cluster,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 

fit2 = gee(Symptomatic ~ contact_with_symptomatic + contact_with_superspreader +
             Age + Gender_FM + Comorbidity + HCW + Hospital_cluster,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 
```

In a multivariate analysis, being a superspreader slightly increased the risk of developing a symptomatic form (OR `r geeORtable(fit1)[grepl("superspreader", geeORtable(fit1)[,1]),2]`, p= `r geeORtable(fit1)[grepl("superspreader", geeORtable(fit1)[,1]),3]`), and adding this variable didn't modify the relationship between symptomatic form in infectee and in infector: the OR of being a contact of a symptomatic case was `r geeORtable(fit_ref)[grepl("contact_with", geeORtable(fit_ref)[,1]),2]` when the variable *superspreader* was not in the multivariate analysis, and OR was `r geeORtable(fit1)[grepl("contact_with", geeORtable(fit1)[,1]),2]` when adding it.

```{r ,echo=F}
geeORtable(fit1)
```

Similarly, being a contact of a superspreader was not associated with the risk of developing a symptomatic form (OR `r geeORtable(fit2)[grepl("contact_with_superspreader", geeORtable(fit2)[,1]),2]`, p= `r geeORtable(fit2)[grepl("contact_with_superspreader", geeORtable(fit2)[,1]),3]`), and adding this variable didn't modify the relationship between symptomatic form in infectee and in infector: the OR of being a contact of a symptomatic case was `r geeORtable(fit2)[grepl("contact_with_sym",     geeORtable(fit2)[,1]),2]` when adding the variable *contact with superspreader* to the model.

```{r ,echo=F}
geeORtable(fit2)
```

## Superspreaders and risk of severe form 

```{r, echo=FALSE,include=F}
fit_ref = gee(Severe ~ contact_with_severe + 
                Age + Gender_FM + Comorbidity + HCW + Hospital_cluster,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 

fit1 = gee(Severe ~ contact_with_severe + superspreader +
             Age + Gender_FM + Comorbidity + HCW + Hospital_cluster,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 

fit2 = gee(Severe ~ contact_with_severe + contact_with_superspreader +
             Age + Gender_FM + Comorbidity + HCW + Hospital_cluster,
          data=llr,id=wave,family=binomial,
          corstr="exchangeable") 
```

In a multivariate analysis, being a superspreader didn't change the risk of developing a severe form (OR `r geeORtable(fit1)[grepl("superspreader", geeORtable(fit1)[,1]),2]`, p= `r geeORtable(fit1)[grepl("superspreader", geeORtable(fit1)[,1]),3]`), and adding this variable didn't modify the relationship between severe form in infectee and in infector: the OR of being a contact of a severe case was `r geeORtable(fit_ref)[grepl("contact_with", geeORtable(fit_ref)[,1]),2]` when the variable *superspreader* was not in the multivariate analysis, and OR was `r geeORtable(fit1)[grepl("contact_with", geeORtable(fit1)[,1]),2]` when adding it.

```{r ,echo=F}
geeORtable(fit1)
```

Surprisingly, being a contact of a superspreader was associated with a **reduced** risk of developing a severe form (OR `r geeORtable(fit2)[grepl("contact_with_superspreader", geeORtable(fit2)[,1]),2]`, p= `r geeORtable(fit2)[grepl("contact_with_superspreader", geeORtable(fit2)[,1]),3]`), even if adding this variable didn't modify the relationship between severe form in infectee and in infector: the OR of being a contact of a severe case was `r geeORtable(fit2)[grepl("contact_with_sev", geeORtable(fit2)[,1]),2]` when adding the variable *contact with superspreader* to the model.

```{r ,echo=F}
geeORtable(fit2)
```


> To be discussed: possibly an unmeasured confounding factor? (note that adding `age of contact` to the gee model make it impossible to converge)

# Conclusions

* Severity in infector is associated with severity in infectee
* Symptomatic form in infector is associated with symptomatic form in infectee
* These associations are not (much) modified when trying to add confounding variables, such as :
  + Age of infectee
  + Gender
  + Comorbidities
  + Infector symptomatic at the time of last contact (in a subset where n infectors = 1)
  + Wave (clustering variable)
  + Age of infector (in a subset where n infectors = 1)
  + Profession of infectee is healthcare worker
  + Healthcare-associated cluster
  + Being a superspreader
  + Being a contact of a superspreader